{\bf Intrusion Detection through Provenance-Based Histograms}

### 2013-11-24 writeup/paper.tex
about to start evaluation

\begin{abstract}
Data-driven intrusion detection is a common system security technique. Here we present a method for generating a statistical model based approach to intrusion detection, using provenance data generated by a provenance-aware modified file system. Our analyses reveal that provenance data can be directly used to detect intrusions, and an in-depth analysis reveals which data is most useful in distinguishing between safe and malicious programs and program executions. Our histogram based approach requires no prior intuition about the types of exploits potentially being run on a system, and identifies suspicious behavior based on deviations from prior normal system behavior.
\end{abstract}

@@

\section{Introduction}
Provenance is metadata that tracks the history of all changes to files. In provenance-aware storage systems like PASS \cite{pass} and PASSv2 \cite{passv2}, the file system automatically tracks all dependencies whenever a file is created, modified, or deleted. These dependencies include the command that was executed to modify the file, the environment of the command, and all input files. The generated provenance forms a directed acyclic graph of typed nodes (e.g. files, processes, pipes) with properties (e.g. name, execution time, pid) and typed edges (e.g. forked, input, versioning).

@@

Applications that require provenance collection typically place great emphasis on data integrity. Given this priority, a natural goal of provenance systems is to detect intrusions on the system. Somayaji et. al. have developed techniques for detecting intrusions based on system call sequences, and counteract these by exponentially delaying or aborting system calls, rendering the system useless for a malicious attacker \cite{somayaji}\cite{somayaji-recent}. These intrusions include exploiting vulnerabilities in the SSH daemon and sendmail to obtain a shell with root privileges. Somayaji's work relies on building a {\em normal} profile of a process, and then detecting when the process deviates from normal.

Given the extensive amount of data PASS collects on file-file, file-process, and process-process dependencies, it seems plausible to detect intrusions based on provenance data collected by PASS. Tariq et. al. generalize Somayaji's sequences of system calls to sequences of provenance system events \cite{correlated-anomalies} to develop an intrusion detection system (IDS) in a distributed system. Their system uses bloom filters to store hashes of $k$-tuples--representing a sequence of provenance events--and then use these bit vectors to correlate anomalies across multiple hosts. King et. al. have developed Backtracker \cite{backtracker}, a modified Linux kernel that tracks dependencies between operating system objects (files, processes, file names) in a similar fashion to PASS. Given an intrusion, Backtracker builds a backwards casual graph to determine the entry point of an intrusion, and they have extended Backtracker with forward causal graphs to determine possibly tainted files, processes, and hosts from an intrusion in a distributed system \cite{multihost}. However, their system does not find a way of detecting an intrusion. Lei et. al. build similar process-file dependency trees, which they call access trees, and define a compatibility function to compare access trees with the goal of predicting future file accesses based on the current access tree pattern \cite{fileprefetch}.

@@

In this paper, we present a technique to analyze existing provenance data to detect if an intrusion occurred.

In particular, we generate statistical models of normal behavior, against which we can test a node of interest. These statiscal models are histograms of different graph statistics, and we use these histograms to generate kernel density estimations to evaluate a test node. The contributions of this paper are:
\begin{enumerate}
\item A histogram-based technique for using provenance data to detect intrustions
\item An analysis of the different intrusions that can be identified, and the types of provenance which best contribute to these detections
\item A cogent plan for future work for using provenance data to detect intrusions
\end{enumerate}

\section{Background and Related Works}
\subsection{PASS}
Provenance data is meta-data detailing the creation, transformation, and flow of data. PASS \cite{pass, passv2} is a provenance-aware storage system, designed to generate and track this provenance data seamlessly. The PASS system automatically creates and updates provenance data fore every file that was created or modified the volume. The PASS provenance data is very rich data. For example, the provenance for a data file that sent in a zip file would reveal both the name of the program that unzipped the file, as well as the web browser through which the zip file was downloaded! It is our belief that this level of specificity will lend itself to detecting patterns in program usage.

The PASS system runs with minimal time overhead, and a reasonable space overhead (XXX cite real numbers).
\subsection{Intrusion Detection}
Intrusion detection in live systems is a common problem in the systems community. There are two primary components to an intrusion detection system (IDS): (1) the data used to fuel the system's analysis, and (2) the process in which this data is used.

Somayaji et al. \cite{somayaji} aimed to detect intrusions by keeping track of all systems calls that occur during a process's execution time. After collecting this data over a long period of "normal" time, if a proces exhibits system call patterns that deviate from the "normal" patterns for the process, their system flags it as potentialy dangerous, and slows its execution. In this way, a program that continually deviates from normal behavior will be descheduled entirely.

Tariq et al. \cite{correlated-anomalies} extend the work of Somayaji and others, designing a system to run an IDS on a set of distributed nodes. They use the IDS of other researchers, distilling a programs execution into a set of values (a "k-tuple") used to identify information of interest regarding the program's exeuction. Our system is designed to run offline, so we run our algorithms on the full set of provenance data. While this approach is slower, it allows us to take in the full picture of what occurred during a process's execution, without having to preselect the information we think will be useful. This approach is especially helpful because different information leads to the detection of different types of intrusions.

Backtracker \cite{backtracker} is a system most similar to ours in design. It, like PASS, collects data while the system is running, and like our system analyzes it offline for potential intrusions. Whereas PASS collects data directly at the operating system level (using a modified file system), Backtracker runs its target OS inside of a virtual machine, and uses a modified virtual machine monitor to collect data. The data they collect is very similar to the data PASS generates, and their exploits and analysis techniques directly inspired our work. 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% DESIGN
%

\section{Design and Implementation}
Here we present our intrusion detection technique: a general and extendable statistical method for analyzing and comparing provenance data. At a high level, we process a provenance data graph, generating a model of "normal" behavior for each process/file in the graph. Using this model, we can determine the likelihood that a given "test" node is normal or whether it deviates from normal behavior and is thus potentially an exploited or malicious program.
\subsection{Models}
Our approach requires statistical models to evaluate the normalcy of a test node. The statistical model we chose to use was a kernel density estimation (KDE), which estimates the probability density function of a value with unknown distribution. To create the KDE, we needed to provide counts of a particular statistic. In this way, for each statistic that we chose to collect, we would be able to evaluate a test node's normalcy according to that statistic. For example, we could then ask "Does node $X_n$ have a statistically believable number of output processes compared to the other $X$ nodes?"
\subsection{Provenance Statistics}
As described above, in order to generate a KDE model, we need some statistic to measure. As we expected intrusions to differ in terms of the statistics that would stand out, we needed to generate a large number of statistics, and a model for each. With regards to provenance nodes directly, we compiled the following statistics:
\begin{enumerate}
\item The number of input files to the node.
\item The number of input processes to the node.
\item The number of input pipes to the node.
\item The number of output files to the node.
\item The number of output processes to the node.
\item The number of ouput pipes to the node.
\item file\_in\_version The "version" number of an incoming file.
\item file\_out\_version The "version" number of an outgoing file.
\end{enumerate}
For ease of computation and comparison, we treated the different node types (e.g. file, process, pipe) separately so that we actually had different counts for "input files to a file node" and "input files to a process node", for instance.
\subsection{Centrality Metrics}
In addition to the data that can be collected on the nodes directly, we could also compute clustering statistics by considering each node in the context of the larger graph. Using the methods espoused by Macko et al. \cite{clustering}, we could compute different centrality and clustering metrics for a node, which can then be tabulated into a histogram like the input/output counts above. 
KENNY TODO: EXPAND THIS
\subsection{Implementation}
The implementation of the above techniques consists of multiple Python files, spanning about 850 lines of (commented) code. The raw BerkeleyDB database generated by PASS is provided as input, and textual/graphical results are outputted.
KENNY TODO: EXPAND THIS
\subsection{Examples}
In Figure 1, we see a few examples of a generated histogram and KDE model. The green bars represent the histogram for the particular statistic, and the blue curve is the KDE. Finally, the red line is the value for a "test" node. In this case, the test node appears to be a possible exploit, because it intersects the KDE at low values. 
\begin{figure*}
  \caption{Example histograms of a centrality measure and process statistic, with the assoicated KDE}
  \centering
    \includegraphics[width=\textwidth]{img/hist.png}
\end{figure*}

>>>>

\begin{abstract}
Data-driven intrusion detection is a common system security technique. We present a histogram-based approach to intrusion detection, using provenance data generated by a provenance-aware storage system. By looking at simple provenance statistics and centrality metrics on provenance graphs, our approach identifies suspicious behavior based on deviations from prior normal system behavior. Our histogram-based approach requires no prior intuition about the types of exploits potentially being run on a system, and our analysis of two buffer overflow attacks and a modified version of \texttt{gcc} to log actions reveal which data is most useful in distinguishing between safe and malicious programs and program executions.
\end{abstract}

@@

\section{Introduction}
Provenance is metadata that tracks the history of all changes to files. In provenance-aware storage systems like PASS \cite{pass} and PASSv2 \cite{passv2}, the file system automatically tracks all dependencies whenever a file is created, modified, or deleted. These dependencies include the command that was executed to modify the file, the environment of the command, and all input files. The generated provenance forms a directed acyclic graph of typed nodes (e.g. files, processes, pipes) with properties (e.g. name, execution time, pid) and typed edges (e.g. forked, input, versioning). The edges show dependency, not data flow, and as a result, edges are directed from outputs to inputs.

@@

Applications that require provenance collection typically place great emphasis on data integrity. Given this priority, a natural goal of provenance systems is to detect intrusions on the system. Somayaji et. al. have developed techniques for detecting intrusions based on system call sequences, and their system counteracts these intrusions by exponentially delaying or aborting system calls, rendering the system useless for a malicious attacker \cite{somayaji}\cite{somayaji-recent}. These intrusions include exploiting vulnerabilities in the SSH daemon and sendmail to obtain a shell with root privileges. Somayaji's work relies on building a {\em normal} profile of a process, and then detecting when the process deviates from normal.

Tariq et. al. generalize Somayaji's sequences of system calls to sequences of system events \cite{correlated-anomalies} to develop an intrusion detection system (IDS) in a distributed system. Their system uses bloom filters to store hashes of $k$-tuples--representing a sequence of systemevents--and then use these bit vectors to correlate anomalies across multiple hosts back to the associated provenance. King et. al. have developed Backtracker \cite{backtracker}, a modified Linux kernel that tracks dependencies between operating system objects (files, processes, file names) in a similar fashion to PASS. Given an intrusion, Backtracker builds a backwards casual graph to determine the entry point of an intrusion, and they have extended Backtracker with forward causal graphs to determine possibly tainted files, processes, and hosts from an intrusion in a distributed system \cite{multihost}. However, their system does not have a way of detecting an intrusion. Lei et. al. build similar process-file dependency trees, which they call access trees, and define a compatibility function to compare access trees with the goal of predicting future file accesses based on the current access tree pattern \cite{fileprefetch}.

@@

Given the extensive amount of data PASS collects on file-file, file-process, and process-process dependencies and existing work in intrusion detection that uses these kinds of directed graphs of files and process data, it seems plausible to detect intrusions based on provenance data collected by PASS. In this paper, we present a histogram-based technique for analyzing existing PASS provenance data to detect if an intrusion occurred.

In particular, we generate statistical models of normal behavior, against which we can test a node of interest. These statistical models are histograms of different graph statistics, and we use these histograms to generate kernel density estimations to evaluate a test node. The lower the kernel density estimate, the less ``normal" the node. The contributions of this paper are:
\begin{enumerate}
\item A histogram-based technique for using provenance data to detect intrusions
\item An analysis of the different intrusions that can be identified, and the types of provenance which best contribute to these detections
\item Future extensions to our work to create better intrusion detection models and make more use of provenance data
\end{enumerate}

\section{Background and Related Works}
\subsection{PASS}
Provenance data is met-data detailing the creation, transformation, and flow of data. PASS \cite{pass, passv2} is a provenance-aware storage system, designed to generate and track this provenance data seamlessly, and Table \ref{passobjects} indicates the types of objects the PASS tracks and attributes with those objects. PASS allocates pnode (``PASS" node) number and version numbers whenever PASS encounters a new object or a modification to an existing object. The combination of (pnode number, version number) uniquely identifies a node in our provenance graph.

The PASS system automatically creates and updates provenance data for every file that was created or modified the volume. For example, the provenance for a data file that sent in a zip file would reveal both the name of the program that unzipped the file, as well as the web browser through which the zip file was downloaded! It is our belief that this level of specificity will lend itself to detecting patterns in program usage.

There are two versions of PASS. We conducted our experiments only with PASSv2; thus in this paper ``PASS" to refer to PASSv2.

TODO: The PASS system runs with minimal time overhead, and a reasonable space overhead (XXX cite real numbers).

\begin{table}[htdp]
\begin{center}
\begin{tabular}{|c|c|}
\hline
{\bf Node Types} & {\bf Description} \\
\hline
PROC & process \\
FILE & file \\
PIPE & pipe \\
NPFILE & non-PASS file \\
OTHER & other types of objects \\
\hline
{\bf Node Attributes} & {\bf Description} \\
\hline
INODE & inode of file \\
PATH & path of file \\
FREEZETIME & timestamp of record \\
CREAT & creation of file \\
UNLINK & unlinking file \\
EXECTIME & exec time of process\\
NAME & name of process \\
PID & pid of of process \\
ARGV & argument vector to process \\
ENV & environment of a process \\
\hline
{\bf Edge Types} & {\bf Description} \\
\hline
FORKPARENT & indicates parent PROC\\
&  of existing PROC node \\
INPUT & indicates input files\\
& or processes \\
VERSION & indicates a modification  \\
& to an existing node (e.g. \\
& modifying a file) \\
\hline
\end{tabular}
\end{center}
\label{passobjects}
\caption{Types of data collected by PASS}
\end{table}%

\subsection{Intrusion Detection}
Intrusion detection in live systems is a common problem in the systems community. There are two primary components to an intrusion detection system (IDS): (1) the data used to fuel the system's analysis, and (2) how the data is used.

Somayaji et al. \cite{somayaji} aim to detect intrusions by keeping track of all systems calls that occur during a process's execution time. After collecting this data over a long period of "normal" time, if a proces exhibits system call patterns that deviate from the ``normal" patterns for the process, their system flags it as potentially dangerous, and slows its execution. In this way, a program that continually deviates from normal behavior will be descheduled entirely.

Tariq et al. \cite{correlated-anomalies} extend the work of Somayaji, designing a system to run an IDS on a set of distributed nodes. They use the IDS of other researchers, distilling a programs execution into a set of $k$-tuples of system events used to identify normal behavior during the program's execution. Our system is designed to run offline, so we run our algorithms on the full set of provenance data. While this approach is slower, it allows us to take in the full picture of the events during a process's execution, without having to preselect the information we think will be useful. This approach is especially helpful because different information leads to the detection of different types of intrusions.

Backtracker \cite{backtracker} is a system most similar to ours in design. It, like PASS, builds a directed graph of process-file dependencies while the system is running, and, like our system, analyzes it offline for potential intrusions. The data they collect is very similar to the data PASS generates, and their exploits and analysis techniques directly inspired our work.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% DESIGN
%

\section{Design and Implementation}
Here we present our intrusion detection technique: a general and extendable statistical method for analyzing and comparing provenance data. At a high level, we create a provenance data graph and generate a model of ``normal" behavior for each process/file in the graph. Using this model, we can determine the likelihood that a given ``test" deviates from normal behavior and is thus potentially an exploited or malicious program.
\subsection{Models}
Our approach requires statistical models to evaluate the normalcy of a test node. The statistical model we chose to use was a kernel density estimation (KDE), which estimates the probability density function of a value with unknown distribution. To create the KDE, we needed to provide counts of a particular statistic. For each statistic:
\begin{enumerate}
\item We computed the counts for that statistic over all the nodes in the provenance graph.
\item We aggregated the counts by the NAME field of the node (e.g. the path to a file for a FILE node, or path to an executable for PROC node). For objects with no NAME field (e.g. PIPE objects), we used the tuple of (pnode number, VERSION number) as the node's name.
\item For each unique name in the graph, we use the counts on that name to generate a histogram of ``normal" behavior. We convert these histograms into kernel density objects to smoothen our histogram and generalize to all continuous values for the statistic.
\end{enumerate}

This model allows us to evaluate a test node's normalcy according to that statistic. For example, we could then ask ``Does node $X'$ have a statistically believable number of output processes compared to the other $X$ nodes?"

\subsection{Provenance Statistics}
We expected intrusions to differ in terms of the statistics that would stand out, and as a result, we needed to generate a diverse suite of statistics and a model for each. For our simple provenance graph statistics, we compiled the following statistics:
\begin{enumerate}
\item The number of input files to the node.
\item The number of input processes to the node.
\item The number of input pipes to the node.
\item The number of output files to the node.
\item The number of output processes to the node.
\item The number of output pipes to the node.
\item The number of VERSION edges into this node.
\item The number of VERSION edges out of this node.
\end{enumerate}
For ease of computation and comparison, we treated the different node types (e.g. file, process, pipe) separately so that we can treat, for instance, "input files to a file node" and "input files to a process node" as separate counts.

\subsection{Centrality Metrics}
In addition to the data that can be collected on the nodes directly, we computed centrality metrics on the nodes using the centrality metrics proposed by Macko et. al. \cite{clustering}. In their work, they noted 3 properties of provenance graphs:
\begin{enumerate}
\item The ubiquity (importance) of a node is a function of a node's descendants, not its ancestors. (Note: Because edges in provenance graphs are in the direction of data dependence instead of data flow, ``descendants" in this case means data flow descendants and NOT descendants by traversing provenance graph edges).
\item There is no agreed-upon granularity in which provenance should be captured.
\item Provenance has a temporal component.
\end{enumerate}

Using these observations, they propose several centrality metrics on provenance graphs:
\begin{itemize}
\item \textbf{In-degree Centrality}. Because a node's importance is based on its output descendants, then the number of times a node was used as input to another node indicates a simple measure of importance.
\item \textbf{Age}. They note that large "jumps" in timestamps or age might indicate breaks between tasks, but this interpretation is complicated when scripts execute tasks in short succession.
\item \textbf{Ancestor Centrality}. This measures the total number of descendants of a node, normalized by the total number of nodes in a graph.
\item \textbf{Opsahl's Closeness Centrality}. This accounts for the number of edges between a node and its descendants. To compute this, we calculate
$$CC'(v) = \sum_{x \in V -  \{v\}} (d'_{xv})^{-1}$$
where $d'_{xv}$ is the distance from node $x$ to $v$ by only following outgoing edges from $x$. Unreachable nodes from $x$ will have $d'_{xv} = \infty$, and so $(d'_{xv})^{-1} = 0$.
\item \textbf{Provenance Eigenvector Centrality}. The centrality of the nodes is given by the left dominant eigenvector of the matrix:
$$M_{ij} =
\begin{cases}
1 & \text{if there is an edge } i \to j \\
1/|V| & \text{if there is no out edge from } i \\
0 & \text{otherwise}
\end{cases}
$$
\end{itemize}

We each metric, we computed the centrality of all the nodes and aggregated the counts by name to generate our histogram and kernel density estimations like the simple statistics we used above.

\subsection{Implementation}
We used Orbiter \cite{orbiter} to visualize and examine the provenance graphs. We wrote Python scripts to (1) parse the raw BerkeleyDB database files storing the PASS data (2) build the provenance graphs (3) collect the counts for each statistic (which include running the centrality calculations), and (4) optionally outputting the histograms and kernel density estimations for a node of interest in a second provenance graph.

\subsection{Examples}
In Figure 1, we see a few examples of a generated histogram and KDE model. The green bars represent the histogram for the particular statistic, and the blue curve is the KDE. Finally, the red line is the value for a ``test" node. In this case, the test node appears to be a possible exploit, because it intersects the KDE at low values. 
\begin{figure*}
  \caption{Example histograms of a centrality measure and process statistic for the \texttt{gcc} exploit, with the associated KDE. The green bars represent the raw histogram counts, the blue curve represents the kernel density function, and the red vertical line represents the estimation of a node of interest.}
  \centering
    \includegraphics[width=\textwidth]{img/hist.png}
\end{figure*}
